---
title: "Medical Terminology Chronology"
author: "Jackson M Luckey"
date: "5/13/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loadPackages}
library(tidyverse)
library(tidynlm)
```

## National Library of Medicine

## tidynlm

I recently wrote the R package tidynlm, which provides easy access to the National Library of Medicine (NLM) API. The package allows the user to create a dataframe of NLM documents matching a particular search query using the function `searchnlm`. The resulting dataframe contains information on attributes such as authorship and date and place of publication. This enables easy analysis of historical medical documents both within R and with other tools such as Palladio. The documents themselves can be accessed with the urls in the url column of the resulting dataframe. In this project, I will be using tidynlm, R, and the tools we learned in Digital Humanities to analyze the history of opioid medical terminology.

## Opioids

```{r opioidData}
term_df <- function(term) {
  tmp <- searchnlm(term, retmax = 10000, output = "wide", collapse_to_first = TRUE)
  tmp$term <- term
  tmp
}
terms <- c("opioid", "opium", "heroin", "poppy", "opiate", "narcotic", "fentanyl", "morphine", "diacetylmorphine", "laudanum")
opioids <- map_dfr(terms, term_df)
```

Opioids are a group of powerful analgesic drugs [@kastinHandbookBiologicallyActive2013]. They both alleviate pain and create intense euphoria, and are considered highly addictive [@macyDopesickDealersDoctors2018; @quinonesDreamlandTrueTale2016]. Opioids generate these effects via activating opioid receptors [@kastinHandbookBiologicallyActive2013]. Opium--the original opioid--is processed liquid from opium poppies, and has been used for pleasure, pain alleviation, and controlling dysentery for thousands of years. In the early 1800s, opium was first purified into morphine, which was later transformed into heroin in the 1870s. More opium derivatives, as well as fully synthetic opioids, were developed in the 1900s. Today, the family of opioids includes heroin, morphine, Fentanyl, oxycodone, hydrocodone, buprenorphine, and methadone. Unfortunately, these new stronger opioids increased the risk of overdose, which is when a high dose of opioids causes respiratory failure among other things. While opioids are highly addictive and considered dangerous, they remain uniquely effective at treating severe pain [@brummettManagingPainEssentials2013; @rastegarAmericanSocietyAddiction2015].

### List of Opioid Terminology
```{r, results="asis", echo=FALSE}
# creates a rmarkown list
# "" ensures that the first term is part of the list (due to "\n -")
cat("", terms, sep = "\n - ")
```

### Opioid Terminology Count

```{r opioidTermCount}
opioids %>%
  group_by(term) %>%
  count() %>%
  arrange(desc(n))
```

```{r fixOpioidDate}
opioids <- opioids %>% mutate(date = parse_number(date))
```

### Opioid Terminology Timeseries

```{r opioidTermTimeseries}
opioids %>%
  # filter out uncommon terms -> makes legend more legible
  filter(term %in% c("narcotic", "opiate", "opium", "poppy", "morphine")) %>%
  ggplot(aes(x = date, color = term)) +
  geom_density() +
  theme_classic() +
  xlim(1750, 2020) +
  scale_color_viridis_d() +
  labs(title = "Changes in Opioid Medical Terminology Over Time",
       subtitle = "National Library of Medicine Archives",
       y = "",
       x = "Date",
       color = "Opioid Term") +
  # add history
  geom_vline(xintercept = 1865) + # End of civil war; Start of soldier's disease
  geom_vline(xintercept = 1914) + # Harrison Narcotic Act
  annotate(
    "label",
    x = c(1845, 1930),
    y = 0.02,
    label = c("End of Civil War", "Harrison Narcotic Act")
  )
```